---
title:  "`r config$publication_top_name`"
pagetitle: HRT England monthly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
   $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:relative; float: right; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

  <style type="text/css">

  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 

`r mumhquarterly::infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have changed the methodology around the drugs included as Hormone Replacement Therapy (HRT) items. This release includes all drugs that are covered by the HRT Prescription Prepayment Certificate (PPC), introduced in April 2023. The HRT PPC provides exemption from prescription charges for drugs licensed to treat menopause in the UK.
A full back series has been included in this release to allow users to make appropriate comparisons between time periods.

The previous publication included all drugs in the following British National Formulary (BNF) paragraphs:

* 6.4.1 - female sex hormones and their modulators
* 7.2.1 - preparations for vaginal and vulval changes.

We are interested in any feedback about the publication or these changes, which you can send by using our ", htmltools::HTML("<a href='https://online1.snapsurveys.com/Official_Statistics_Feedback'>Official Statistics feedback survey</a>"),"."), width = "100%")`


## Key findings

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>11 million HRT items were prescribed to an estimated 2.3 million identified patients in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>More than twice as many patients were prescribed HRT items in the least deprived areas compared to the most deprived in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>The proportion of HRT prescribing that is exempt from prescription charges has increased from April 2023.</b>", width = "100%")`
:::
::::
In England in 2022/23:

* there were 11 million items for HRT prescribed in England, a 47% increase from 2021/22

* there were an estimated 2.3 million identified patients that were prescribed HRT items, a 29% increase from  2021/22

* more than twice as many patients were prescribed HRT items in the least deprived areas compared to the most deprived

* the proportion of HRT prescribing that is exempt from prescription charges has increased from April 2023, reversing a downwards trend from 2015/16 onwards. This coincides with the introduction of the HRT PPC

---

## 1. Things you should know


`r mumhquarterly::infoBox_border("Background ", text = "Hormone Replacement Therapy (HRT) is a treatment to relieve symptoms of the menopause. It replaces hormones that are at a lower level during the perimenopause, menopause, and postmenopause.

Menopause is when periods stop due to lower hormone levels. It usually affects women between the ages of 45 and 55, but it can happen earlier.
It affects anyone who has periods. Many women experience some symptoms, such as vaginal dryness, low mood, anxiety, hot flushes and difficulty sleeping. Some of these symptoms can be severe and have a negative impact on everyday life.", width = "100%")`

### 1.1. Introduction {.toc-ignore} 

This publication aims to describe the prescribing of HRT medications in England. These medications are subsequently dispensed in the community in England, Scotland, Wales, Isle of Man or the Channel Islands. Dispensing can be done by a pharmacy, appliance contractor, dispensing doctor, or have been personally administered by a GP practice. They do not include data on medicines used in secondary care, prisons, or issued by a private prescriber. The statistics do not include prescriptions that were issued but not presented for dispensing or where there has been an unmet demand for treatment.

### 1.2. Scope {.toc-ignore} 

These statistics detail:

* the total number of prescription items issued for drugs used in HRT
* the number of identified patients that have received prescribing for these drugs
* demographic breakdowns of prescribing by age group
* demographic breakdowns by a measure of deprivation.

The costs associated with the provision of these medicines are not discussed in this summary. However, they are available within the supporting summary tables that accompany this release.

There is no singular definition of HRT medications. Many other products may be used in the treatment of the menopause that are not included in these statistics.

The release includes all drugs that are covered by the HRT PPC which is intended to cover all drugs licensed to treat menopause in the UK. You can find the most up to date list of drugs included in the HRT PPC on [the NHSBSA website.](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc)

These statistics use the therapeutic classifications defined in the BNF, using the classification system prior to BNF edition 70. Each January the NHSBSA updates the classification of drugs within the BNF hierarchy which may involve some drugs changing classification between years of data. 

Medicines are classified by their primary therapeutic indication. However, it is possible that they can be prescribed for other reasons outside of this primary therapeutic indication. For example, the  treatment of gender dysphoria. The clinical indication of a prescription is not captured by NHSBSA during processing, and therefore we cannot determine the reason that a prescription was issued. 

### 1.3. Definitions {.toc-ignore}

`r mumhquarterly::infoBox_border("Item", text = "A single unit of medication listed separately on a prescription form. In this publication, an example of an item would be Estradiol 2mg tablets x84.", width = "100%")`

`r mumhquarterly::infoBox_border("Patient", text = "A unique NHS number verified from a prescription form. Verification of an NHS number for patient identification is done by the NHS Personal Demographics Service (PDS).", width = "100%")`

`r mumhquarterly::infoBox_border("NIC", text = "The Net Ingredient Cost (NIC) is the basic price of the medication and the quantity prescribed. It does not include other fees incurred by dispensing contractors, such as controlled drug fees or the single activity fee. The basic price is determined by the [Drug Tariff](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff) or by the manufacturer, wholesaler, or supplier of the product.", width = "100%")`

`r mumhquarterly::infoBox_border("Age", text = "Where patients are identified, their age is calculated on 30 September of the given financial year to assign them to a single age band. For patients where date of birth has not been captured, they have been included in a category of unknown.", width = "100%")`





### 1.4. Time periods {.toc-ignore}

This publication presents data from April 2015 to `r ltst_month_tidy`, the most recent available. The data is available in tables at both a monthly and a financial year level. The most recent financial year, 2023/24, shows a year to date view of the figures from April 2023 to June 2023.

### 1.5. Patient identification {.toc-ignore}

When the NHSBSA processes prescriptions, the NHS number of the patient is identified for most, but not all, prescriptions. The proportion of items for which a patient could be identified is shown in Table 1. This means that data relating to patient counts represents most, but not all, patients. 

These statistics do not contain any personally identifiable data. More information about how the NHSBSA protects personal information can be found in the [statement on confidentiality and access](https://www.nhsbsa.nhs.uk/policies-and-procedures).

#### Table 1: The proportion of items for which an NHS number was recorded for listed BNF paragraphs

```{r pat_ident_rates, out.width="100%"}
table_1
```
`r get_download_button(title = "Download data table", data = table_1_data, filename = "table_1")`

---

## 2. Results and commentary

### Figure 1: Number of prescribed items and estimated identified patients by financial year {.toc-ignore}

```{r annual_summary, out.width="100%"}

figure_1

```
`r get_download_button(title = "Download data table", data = figure_1_data, filename = "figure_1")`
Source: [Financial Year Summary Tables - National Totals table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r infoBox_no_border("", text = "<b> There were 11 million HRT items prescribed in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>There were 2.3 million identified patients who received an HRT item in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>Items and patients continue to trend upwards for HRT.</b>", width = "100%")`
:::
::::

There were 10.9 million HRT drug items prescribed in 2022/23. This was a 46.9% increase from 7.44 million items compared with the previous year. 

There were an estimated 2.34 million identified patients who were prescribed at least one HRT item in 2022/23. This was a 29.3% increase from 1.81 million identified patients when compared with 2021/22. 

From April 2023 to June 2023 there were 3.04 million HRT drug items prescribed. This is 597,000 more items than the same period in the previous year, a 24.4% increase.

### Figure 2: Number of prescribed items and estimated identified patients by month {.toc-ignore}

```{r monthly_summary, out.width="100%"}

figure_2
  
```
`r get_download_button(title = "Download data table", data = figure_2_data, filename = "figure_2")`
Source: [Monthly Summary Tables - National Totals table](`r config$month_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 1 million HRT items prescribed in June 2023.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 800,000 identified patients who received an HRT item in June 2023.</b>", width = "100%")`
:::
::::


There were 1.04 million HRT drug items prescribed in June 2023. This was a 0.58% increase from 1.03 million items compared with the previous month, and an increase of 26.1% from 825,000 items in June 2022. 

There were an estimated 797,000 identified patients who were prescribed at least one HRT item in June 2023. This was a 1.76% decrease from 784,000 identified patients when compared with the previous month. 

### Figure 3: Number of prescribed items and estimated identified patients by financial year and selected drug groups {.toc-ignore}

In 2022/23 almost 99% of all HRT items were from the drug groups, female sex hormones and their modulators, and preparations for vaginal and vulval changes. 

These charts show a rolling five years, along with the year to date for readability and the full data is available in the accompanying tables. 

```{r, figures-side, fig.show="hold", out.width="50%"}
 hw_grid(figure_3_pt_1, 
figure_3_pt_2)
```
`r get_download_button(title = "Download data table", data = figure_3_data, filename = "figure_3")`
Source: [Financial Year Summary Tables - Chemical Substance table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 8.3 million female sex hormones and their modulators items prescribed in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 2.5 million preparations for vaginal and vulval changes items prescribed in 2022/23.</b>", width = "100%")`
:::
::::

There were 8.25 million female sex hormones and their modulators items prescribed in 2022/23. This was a 57.0% increase from 5.26 million items compared with the previous year. 

There were an estimated 1.61 million identified patients who were prescribed at least one female sex hormones and their modulators item in 2022/23. This was a 39.8% increase from 1.15 million identified patients when compared with 2021/22. 

There were 2.54 million preparations for vaginal and vulval changes items prescribed in 2022/23. This was a 23.5% increase from 2.05 million items compared with the previous year. 

There were an estimated 833,000 identified patients who were prescribed at least one preparations for vaginal and vulval changes item in 2022/23. This was a 21.6% increase from 685,000 identified patients when compared with 2021/22. 


### Figure 4: Estimated number of identified patients by financial year for selected BNF chemical substances {.toc-ignore}

A chemical substance is the name of the main active ingredient in a drug. The chemical substances featured in figure 4 and 5 are the top 5 most prescribed chemical substances for HRT prescribing. Although chemical substance is a grouping within the BNF, when a drug is available in different presentations with disparate uses it may be listed in multiple sections or paragraphs. In this case it will be assigned multiple chemical substances codes, one for each section it is in. 

This is the case for Estradiol which has a code for female sex hormones and their modulators and preparations for vaginal and vulval changes. The chemical substance code is included in the figures and commentary to clarify between the prescribing for each drug group.

```{r annual_chem_sub, out.width="100%"}

figure_4

```
`r get_download_button(title = "Download data table", data = figure_4_data, filename = "figure_4")`
Source: [Financial Year Summary Tables - Chemical Substance table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> Estradiol (0604011G0) was the chemical substance which had the largest number of estimated identified patients in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 950,000 identified patients who were prescribed at least one Estradiol (0604011G0) item in 2022/23.</b>", width = "100%")`
:::
::::

Between 2021/22 and 2022/23 Estradiol (0604011G0) was the chemical substance which had the largest increase in estimated identified patients who were prescribed at least one item. This was an increase of 61.1% or 362,000 patients from 593,000 in 2021/22 to 955,000 in 2022/23.


### Figure 5: Number of prescribed items by financial year for selected BNF chemical substances {.toc-ignore}

```{r chem_sub_items, out.width="100%"}

figure_5

```
`r get_download_button(title = "Download data table", data = figure_5_data, filename = "figure_5")`
Source: [Financial Year Summary Tables - Chemical Substance table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> Estradiol (0604011G0) was the chemical substance which had the largest number of prescribed items in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 3.9 million prescribed items for Estradiol (0604011G0) in 2022/23.</b>", width = "100%")`
:::
::::

Between 2021/22 and 2022/23 Estradiol (0604011G0) was the chemical substance which had the largest increase in the number of items prescribed. This was an increase of 70.2% or 1.62 million items from 2.30 million in 2021/22 to 3.92 million in 2022/23.

### Table 2: Top 10 presentations by prescribed items in `r ltst_year`{.toc-ignore}

`r mumhquarterly::infoBox_border(header = "Presentations", text = paste0("Pharmaceutical presentations can come in many forms including tablets, capsules, patches, gels, creams and more. Please see the definition of items and quantity given in section 1.2 to help interpret tables 2 and 3. The chemical substance listed in figures 4 and 5 can be made up of more than one of the presentations in tables 2 and 3. For example both Vagifem 10 microgram vaginal tablets and Estradiol 10  microgram pessaries have the chemical substance of Estradiol."), width = "100%")`

Tables 2 and 3 are based on the BNF presentation that has appeared on the prescription form. It is possible that products that are different brands, or generic equivalents, of the same product may appear in these tables. For example Oestrogel Pump-Pack 0.06% gel is the branded equivalent of Estradiol 0.06% gel (750 microgram per actuation) and Vagifem 10 microgram vaginal tablets is the branded equivalent of Estradiol 10 microgram pessaries.

``` {r top_pres_fy, out.width = "100%"}
table_2


```
`r get_download_button(title = "Download data table", data = table_2_data, filename = "table_2")`
Source: [Financial Year Summary Tables - Presentations](`r config$fy_excel_link`)


In 2022/23 Utrogestan 100mg capsules had the most prescribed items with 935,000 total items. This is more than double than the 458,000 items in 2021/22.

### Table 3: Top 10 presentations by prescribed items in `r ltst_month_tidy` {.toc-ignore}

```{r top_pres_month, out.width="100%"}

table_3

```
`r get_download_button(title = "Download data table", data = table_3_data, filename = "table_3")`
Source: [Monthly Summary Tables - Presentations](`r config$month_excel_link`)

In June 2023, Estradiol 0.06% gel had the most prescribed items at 91,900 total items.


### Figure 6: Estimated number of identified patients per 1,000 population by Integrated Care Board (ICB) in `r ltst_year` {.toc-ignore}

```{r patients_per_1000, out.width = "100%"}


figure_6$x$hc_opts$series[[1]]$dataLabels$enabled  <- FALSE

figure_6
```
`r get_download_button(title = "Download data table", data = figure_6_data, filename = "figure_6")`
Source: [Financial Year Summary Tables - ICB Totals table](`r config$fy_excel_link`)


`r mumhquarterly::infoBox_border(header = "ONS Population", text = paste0("The figures in Figure 6 have been calculated by joining the [Office for National Statistics (ONS) Mid-2020 Population Estimates for Lower Layer Super Output Areas in England](https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates/mid2020sape23dt2/sape23dt2mid2020lsoasyoaestimatesunformatted.xlsx) to our data to obtain the estimated population of each ICB."), width = "100%")`

NHS North East London ICB had the lowest estimated number of identified patients who received at least one HRT item per 1,000 residents. It had an estimated 17 patients per 1,000 population in 2022/23. 

NHS Gloucestershire Integrated Care Board was the ICB with the highest estimated number of identified patients who received at least one HRT item per 1,000 residents. It had an estimated 62 patients per 1,000 population in 2022/23.

### Figure 7: Estimated number of identified patients by age band in `r ltst_year` {.toc-ignore}

```{r ageband_fy, out.width="100%"}
figure_7$x$hc_opts$series[[1]]$dataLabels$allowOverlap <- TRUE
figure_7
```
`r get_download_button(title = "Download data table", data = figure_7_data, filename = "figure_7")`
Source: [Financial Year Summary Tables - Age Band table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> Patients aged 50 to 54 were the largest group of identified patients in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 580,000 patients in the 50-54 age band, 24.6% of all patients in 2022/23.</b>", width = "100%")`
:::
::::

The most common group to receive prescribing for HRT in 2022/23 was patients aged 50 to 54. This group had an estimated 580,000 identified patients, 24.6% of all identified patients. The next most common groups were patients aged 55 to 59 with 466,000 and 19.8% of identified patients, and patients aged 45 to 49 with 344,000 and 14.6% of identified patients.

More information on how we calculate a patient’s age can be found in section 4 of this summary.

### Figure 8: Estimated number of identified patients by IMD quintile in `r ltst_year` {.toc-ignore}

``` {r imd_fy, out.width = "100%"}

figure_8$x$hc_opts$series[[1]]$dataLabels$allowOverlap <- TRUE
figure_8
```

`r get_download_button(title = "Download data table", data = figure_8_data, filename = "figure_8")`
Source: [Financial Year Summary Tables - IMD Quintile table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>660,000 patients were prescribed HRT items in the least deprived areas in England in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>More than twice as many patients were prescribed HRT items in the least deprived areas compared to the most deprived in 2022/23.</b>", width = "100%")`
:::
::::


In 2022/23, there were an estimated 662,000 identified patients prescribed HRT medicines in the least deprived areas in England. This was more than double the 289,000 identified patients that received prescribing from in the more deprived areas. In general, more people were prescribed HRT medicines in the least deprived areas in 2022/23. This pattern has remained consistent since 2015/16.

The English Indices of Deprivation have been used to provide a measure of patient deprivation. The deprivation quintiles reported are those of the patient's registered address. More information on this measure can be found in section 4 of this summary.

### Figure 9: Proportion of patients under/over 60 by IMD quintile in `r ltst_year` {.toc-ignore}

``` {r imd_age, out.width = "100%"}

figure_9
```
`r get_download_button(title = "Download data table", data = figure_9_data, filename = "figure_9")`
Source: [Financial Year Summary Tables - IMD Quintile Age table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> There were 215,000 patients under 60 in the most deprived areas in England in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 439,000 patients under 60 in the least deprived areas in England in 2022/23.</b>", width = "100%")`
:::
::::


In 2022/23 there were an estimated 215,000 identified patients under the age of 60 in the most deprived areas who were prescribed at least one HRT item. This was 74.8% of the total identified patients in the most deprived areas. In the least deprived areas 439,000 identified patients were under 60. This was 66.6% of the total identified patients with a captured age in the least deprived area who were prescribed at least one HRT item.

### Figure 10: Proportion of chargeable items by financial year {.toc-ignore}

``` {r charge_fy, out.width = "100%"}

figure_10

```
`r get_download_button(title = "Download data table", data = figure_10_data, filename = "figure_10")`
Source: [Financial Year Summary Tables - Exemption Categories table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> 70% of all prescribing for HRT items was exempt from prescription charges in 2022/23.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The proportion of HRT prescribing that is exempt from prescription charges increased in 2023/24 YTD after declining for the previous two years. This coincides with the introduction of the HRT PPC.</b>", width = "100%")`
:::
::::

`r mumhquarterly::infoBox_border(header = "Prescription charges", text = paste0("In England, most working-age adults have to pay prescription charges and any items dispensed to this group are counted as chargeable. Certain groups are exempt from paying for their NHS prescriptions, for example those over the age of 60. Some prescribed items are always free, including contraceptives. 

If a patient regularly pays prescription charges, a Prescription Prepayment certificate (PPC) could save them money by paying one up front cost annually or for a 3 month period. 

An annual HRT PPC was introduced in April 2023, that covers medicines licensed to treat the menopause in the UK. 

Prescriptions covered by either a PPC or HRT PPC are classed as exempt in the data because the patient does not pay when collecting the medication.

It is not possible to break exempt prescriptions down by category,  as this data is not always recorded. More information can be found in section 3 of this summary.
"), width = "100%")`


For HRT items 70.0% were exempt from prescription charges in 2022/23. This was a decrease from 71.3% in  the previous year. This trend has reversed in the first quarter of 2023/24 where 80.9% of HRT prescribed items were exempt from charge. This coincides with the introduction of the HRT PPC.


### Figure 11: Proportion of chargeable items by month April 2022 to June 2023 {.toc-ignore}

``` {r exempt_monthly, out.width = "100%"}

figure_11

```
`r get_download_button(title = "Download data table", data = figure_11_data, filename = "figure_11")`
Source: [Financial Year Summary Tables - Exemption Categories table](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The proportion of HRT prescribing that is exempt from prescription charges increased by 8 percentage points in April 2023. This coincides with the introduction of the HRT PPC.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The proportion of HRT prescribing that is exempt from prescription charges trended upwards in each month of 2023/24 YTD.</b>", width = "100%")`
:::
::::



## 3. Background

The menopause is a biological stage in a woman's life that occurs when she stops menstruating and reaches the end of her natural reproductive life. This is not usually abrupt, but a gradual process during which women experience perimenopause before reaching postmenopause. Usually the menopause is defined as having occurred when a woman has not had a period for 12 consecutive months (for women reaching menopause naturally).

As well as a change in their menstrual cycle women may experience a variety of symptoms associated with menopause such as hot flushes and sweats, low mood, vaginal dryness, low sexual desire, and others.

Most women will experience some symptoms. Symptoms can last for months or years, and can change with time. Prolonged lack of the hormone oestrogen affects the bones and cardiovascular system, and postmenopausal women are at increased risk of a number of long-term conditions, such as osteoporosis.

HRT, which replaces the hormones that are at low levels, is the main medicine treatment for menopause and perimenopause symptoms. More information on the menopause is available on the [NHS website](https://www.nhs.uk/conditions/menopause/). 

### 3.1. Included drugs {.toc-ignore}

This publication includes all drugs that have been covered by a HRT PPC. The HRT PPC exempts patients from charges for drugs licensed to treat menopause in the UK. You can find the most up to date list of drugs included in the HRT PPC on [the NHSBSA website.](https://www.nhsbsa.nhs.uk/help-nhs-prescription-costs/nhs-hormone-replacement-therapy-prescription-prepayment-certificate-hrt-ppc)

A list of all drugs that are included in this release, including any that are no longer listed due to classification changes or discontinuation can be found in [Appendix A](`r config$appendix_link`)

#### 3.1.1. Female sex hormones and their modulators {.toc-ignore}

Oestrogen-only HRT is usually only recommended for women who have had their womb removed during a hysterectomy. For women with a uterus, oestrogen plus progestogen HRT is recommended. A progestogen should be prescribed to reduce the risk of cystic hyperplasia of the endometrium, a thickening of the endometrium, and possible transformation to cancer. 

HRT comes in several different forms such as tablets and skin patches. The treatment routine for HRT depends on whether the patient is in the early stages of the menopause or has had menopausal symptoms for some time. More information on the types and routines of HRT treatment can be found on the [NHS website](https://www.nhs.uk/conditions/hormone-replacement-therapy-hrt/types/).

The 2 types of routines are cyclical (or sequential) HRT and continuous combined HRT.

Continuous combined preparations are not suitable for use in the perimenopause or within 12 months of the last menstrual period. Women who use such preparations may bleed irregularly in the early stages of treatment.

You can find out more information on [sex hormones](https://bnf.nice.org.uk/treatment-summaries/sex-hormones/) on the NICE website.

#### 3.1.2. Preparations for vaginal and vulval changes {.toc-ignore}

Preparations for vaginal and vulval changes are used to treat the symptoms of vaginal atrophy related to oestrogen deficiencies related to the menopause. 

There are non-hormonal vaginal moisturisers available. Menopausal women with vaginal dryness can use vaginal moisturisers and lubricants alone or in addition to vaginal oestrogen.

You can find out more information on [vaginal and vulval conditions](https://bnf.nice.org.uk/treatment-summaries/vaginal-and-vulval-conditions/) on the NICE website.

#### 3.1.3. Anabolic steroids {.toc-ignore}

#### Prasterone

Prasterone is biochemically and biologically identical to endogenous dehydroepiandrosterone (DHEA), and is converted to oestrogens and androgens.

It is used to treat vulvar and vaginal atrophy in postmenopausal women.

You can find out more information on [prasterone](https://bnf.nice.org.uk/drugs/prasterone/) on the NICE website.

#### 3.1.4. Progesterone only contraception {.toc-ignore}

#### Intrauterine system (IUS), or Mirena coil

For patients who have a womb and are taking or using oestrogen tablets, patches, gel or spray, the Mirena coil, an intrauterine system (IUS), may be a suitable way to deliver the progestogen they need.

The Mirena coil is inserted into the womb and it gradually releases a form of progestogen (levonorgestrel) into the body.

You can find out more information on [levonorgestrel](https://bnf.nice.org.uk/drugs/levonorgestrel/) on the NICE website.

### 3.2. Gina 10 microgram vaginal tablets {.toc-ignore}

Pharmacists in the UK can now dispense the HRT drug Gina 10 microgram vaginal tablets without a prescription.

Gina 10 microgram vaginal tablets are a locally acting oestrogen HRT and the only treatment available without a prescription. It has been clinically proven to target and treat the cause of vaginal atrophy and its associated symptoms after the menopause.

Gina 10 microgram vaginal tablets are a proprietary equivalent of Estradiol 10 microgram pessaries. The NHSBSA does not collect information about medicines that are sold over the counter. These statistics therefore may not be representative of the whole population taking HRT.

### 3.3. Shortages {.toc-ignore}

There have been concerns that some women across the UK were unable to obtain their prescriptions due to shortages of HRT products. This coupled with the impact of COVID-related global supply problems means that many women are extremely concerned about not being able to get access to this vital medication.

There are over 70 HRT products available in the UK and while most remain in good supply, a range of factors including an increase in demand has led to supply issues with a limited number of products. The Department of Health and Social Care (DHSC) continues to work closely with suppliers and other stakeholders to maintain overall HRT supply to patients across the UK. This includes working to expedite resupply dates of the disrupted HRT products to resolve issues as soon as possible and to help fill supply gaps and prevent future shortages. Access to in-demand products has improved since Serious Shortage Protocols (SSPs) were issued to limit dispensing to three months’ supply to even out distribution and allow alternative products to be dispensed, as necessary. Suppliers have taken swift action to secure additional stock of HRT products which are experiencing shortages and deliveries of further stock have been expedited. These actions have increased near-term supply and the supply position has improved. To support continued growth in the longer term, suppliers are taking steps to further build capacity and increase production. 

### 3.4. Exemption categories {.toc-ignore}

Exemption category data is not included in these statistics due to a number of known data quality issues.

There are many possible exemption categories that may apply to a specific prescription but only one can be recorded in our data. This is the case even if more than one was indicated by marks on the prescription form.

In a large number of cases the NHSBSA does not record a specific exemption category. A common and increasing reason for our records stating ‘No Declaration/Declaration Not Specific’ is the increasing use of the Real Time Exemption Checking service (RTEC). This service confirms to the dispenser that a person was exempt from the charge, but no record of the reason found by RTEC is passed on to NHSBSA prescription processing systems or data warehouse.

In many cases the mark(s) on the rear of the form intended to indicate one, or more, exemption reason(s) is not clear and can be misinterpreted by the NHSBSA scanners.

NHSBSA performs no validation or verification of which exemption category has been marked or recorded during scanning.

Exemption category data is released alongside this publication to complement the HRT statistics, but is management information only and is not an official statistic. 


## 4. About these statistics

Though this publication is an official statistics in development publication by the NHSBSA. To learn more about how this publication compares to our other publications, you can view the Official Statistics guidance table on our [statistical collections page](https://www.nhsbsa.nhs.uk/statistical-collections). 

Further information on the methodology used in this publication and further background information is available in our [Background Information and Methodology](`r config$background_link`) supporting document.

### 4.1. Changes to these statistics {.toc-ignore}

For this release, we have changed the methodology around the drugs included as HRT items. This release includes all drugs that are covered by the HRT PPC, introduced in April 2023. The HRT PPC provides exemption from prescription charges for drugs licensed to treat menopause in the UK.

The previous publication included all drugs in the following BNF paragraphs:

* 6.4.1 - female sex hormones and their modulators
* 7.2.1 - preparations for vaginal and vulval changes.

### 4.2. Patient counts {.toc-ignore}

The patient counts shown in these statistics should only be analysed at the level at which they are presented. Adding together any patient counts is likely to result in an overestimate of the number of patients. A patient will be included, or counted, in each category or time period in which they received relevant prescriptions. For example, if a patient received a prescription item for HRT in January 2021, and another in February 2021, then adding together those totals would count that patient twice. For the same reason, data on patient counts for different BNF sections should not be added together.

### 4.3. Patient age {.toc-ignore}

The age of patients used in these statistics is derived from data provided by the NHS Personal Demographics Service (PDS) for NHS numbers that have been successfully verified by them. A patient’s age, used to assign them to an age group, has been calculated on 30 September for the given financial year. It is possible that a patient’s PDS information may change over the course of the year. In these cases patients may be subject to multiple counting in these analyses.

### 4.4. Averages {.toc-ignore}

Where this document refers to 'average', this is the mean unless otherwise stated. This is calculated by adding the number of items/patients/months together and dividing this by the number of items/patients/months.

### 4.5. Index of deprivation {.toc-ignore}

The English Indices of Deprivation 2019 have been used to provide a measure of patient deprivation. The English Indices of Deprivation are an official national measure of deprivation that follows an established methodological framework to capture a wide range of individuals’ living conditions.

The reported  Index of Multiple Deprivation (IMD) quintile, where 1 is the most deprived and 5 the least deprived, is derived from the lower-layer super output areas (LSOA) of the registered address of the patient who received the medication. There are a number of items each year that we have reported as having an unknown IMD quintile. These are items that have been attributed to an patient who exclusively received their prescriptions via paper prescribing and not via the Electronic Prescription Service (EPS). LSOA is only captured for prescriptions via EPS. For those patients who received both paper prescribing and EPS we have imputed their LSOA from the latest available data. The use of EPS increased considerably during 2020/21. Therefore historical data has much higher rates of patients with a LSOA which has been reported as 'Unknown'.

IMD quintiles are calculated by ranking census LSOA from most deprived to least deprived and dividing them into 5 equal groups. These range from the most deprived 20% (quintile 1) of small areas nationally to the least deprived 20% (quintile 5) of small areas nationally.

---

## 5. Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: accessibility@nhsbsa.nhs.uk

**Phone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We’ll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Contact us’ section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---

## 6. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics. 

You can complete a [short survey about this publication](`r config$stats_survey_link`) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:** statistics@nhsbsa.nhs.uk

**You can also write to us at:**

NHSBSA - Statistics  
NHS Business Services Authority  
Stella House  
Goldcrest Way  
Newburn Riverside  
Newcastle upon Tyne  
NE15 8NY

**Responsible statistician:** `r config$responsible_statistician`

</main>